---
title: "ANA 515 Assignment 3 Getting and Cleaning Data"
author: "Akshay Patalay"
date: "2024-10-13"
output:
  html_document:
    df_print: paged
---

## File Name : survey_sheet1.csv, survey_sheet2.csv
## 1. Installing packages and libraries into R

Installing required packages and libraries to finish up with the project in order to use the required functions to perform operations.

```{r library, echo=TRUE}
library(tidyverse)
library(knitr)
library(dplyr)
library(tools)
library(stringr)
library(ggplot2)
library(rlang)
```

## 2 Reading the data into R

Used the read.csv function to load the csv file into a data frame. This function to read the csv file has come from the tidyverse package. 

Special options that are available as part of the read.csv function are as follows - 

(1) StringsAsFactors = By default, when building or importing a data frame, the columns that contain characters (i.e., text) are coerced (converted) into the factor data type.
(2) Check Names = Affects how column names are handled when reading CSV files into R.If set to TRUE, it modifies any non-compliant names to conform to R's variable naming rules, such as replacing spaces with dots and converting the first character to lowercase.
(3) Header = If TRUE, read.csv() assumes that your file has a header row so first row is the name of each column.
(4) Separator = The field separator character.
(5) SkipNul = Skip the NULL values from the data set.


```{r read csv, echo=TRUE}
mental_health_sheet1_csv <- read.csv("C:\\Users\\aksha\\Desktop\\ANA515P_01\\survey_sheet1.csv",stringsAsFactors = F,check.names = F, header = T, sep = ",", skipNul = T)
mental_health_sheet1_csv <- as.data.frame(mental_health_sheet1_csv)

mental_health_sheet2_csv <- read.csv("C:\\Users\\aksha\\Desktop\\ANA515P_01\\survey_sheet2.csv",stringsAsFactors = F,check.names = F, header = T, sep = ",", skipNul = T)
mental_health_sheet2_csv <- as.data.frame(mental_health_sheet2_csv)

head(mental_health_sheet1_csv,10)
head(mental_health_sheet2_csv,10)

str(mental_health_sheet1_csv)
str(mental_health_sheet2_csv)

mental_health_sheet2_csv$Age <- as.integer(mental_health_sheet2_csv$Age)

```
## 3 Merging the two data frames into one.

The two files Sheet1 & Sheet2 Survey CSV files are merged together using the merge command.
Printing the values to see total rows from both files are captured in the data frame.

```{r merge data frames, echo=TRUE}
merge_mental_health_df <- merge(x=mental_health_sheet1_csv,y=mental_health_sheet2_csv,all=TRUE)
print(merge_mental_health_df)
```

##  4.	Subset the data frame to the following columns:

Selecting only few columns for analysis and visualizations.

(1) AGE
(2) GENDER
(3) COUNTRY
(4) STATE
(5) FAMILY_HISTORY
(6) TREATMENT
(7) WORK_INTERFERE
(8) NO_EMPLOYEES
(9) REMOTE_WORK
(10) TECH_COMPANY
(11) CARE_OPTIONS
(12) WELLNESS_PROGRAM
(13) LEAVE
(14) COWORKERS
(15) SUPERVISOR


```{r subset columns, echo=TRUE}
myvars <- c("Age", "Gender", "Country", "state","family_history","treatment","work_interfere","no_employees","remote_work","tech_company","care_options","wellness_program","leave","coworkers","supervisor")
merge_mental_health_df <- merge_mental_health_df[myvars]
head(merge_mental_health_df,10)
```

## 5.	Arrange the data in the data frame by STATE

Arranging the data by state in ascending order.

```{r arrange, echo=TRUE}
merge_mental_health_df <- arrange(merge_mental_health_df,state)
head(merge_mental_health_df,10) 
```

## 6.	Limit the data 

Limiting the data to only include :

(1) Population of United States 
(2) considering age group above 15 years of age.

Limiting the data allows to deal or work with a smaller sub group of data for further analysis and visualization

```{r subset, echo=TRUE}
merge_mental_health_df <- subset(merge_mental_health_df, Country == 'United States')

filteredelectricvehicles <- merge_mental_health_df %>%
filter(Age > 15)

print(filteredelectricvehicles)
```

## 7. Summary Statistics and finding unique values per column

(1) Determining the summary statistics i.e min, 1st Q , 3rd Q, max
(2) Finding unique values for each column to determine categories for categorical columns

Summary statistics is a part of descriptive statistics that summarizes and provides the gist of information about the sample data. Statisticians commonly try to describe and characterize the observations by finding: a measure of location, or central tendency, such as the arithmetic mean.

Finding unique values for each column determing unique categories associated with every column.

```{r summary statistics and unique values, echo=TRUE}
summary(filteredelectricvehicles)

unique(filteredelectricvehicles$Age)
unique(filteredelectricvehicles$Gender)
unique(filteredelectricvehicles$Country)
unique(filteredelectricvehicles$state)
unique(filteredelectricvehicles$treatment)
unique(filteredelectricvehicles$family_history)
unique(filteredelectricvehicles$work_interfere)
unique(filteredelectricvehicles$no_employees)
unique(filteredelectricvehicles$remote_work)
unique(filteredelectricvehicles$tech_company)
unique(filteredelectricvehicles$benefits)
unique(filteredelectricvehicles$care_options)
unique(filteredelectricvehicles$wellness_program)
unique(filteredelectricvehicles$leave)
unique(filteredelectricvehicles$coworkers)
unique(filteredelectricvehicles$supervisor)
```
## 8.	Converting all the column names to upper case including spelling 

Convert all the column names to upper case for clarity.

```{r Column Upper Case, echo=TRUE}
subset_storm_events_csv <- rename_all(filteredelectricvehicles,toupper)
print(subset_storm_events_csv)
```

## 9. Formatting the data in the data frames, handling missing values, checking spellings

In the below code we are -

(1) Formatting the data for given columns :
GENDER, FAMILY_HISTORY, TREATMENT, NO_EMPLOYEES, REMOTE_WORK, TECH_COMPANY, CARE_OPTIONS,        WELLNESS_PROGRAM, LEAVE, COWORKERS, SUPERVISOR

To transform and standardize the value for easy analysis and application of machine learning algorithms.

(2) Handling missing values
By assigning NA to missing values

Missing or empty values are handled in the case_when command all missing values are assigned a category as NA.

```{r case statements, echo=TRUE}

subset_storm_events_csv <- subset_storm_events_csv %>% mutate(
                          GENDER = case_when
                              (
                                GENDER %in% c("M") ~ "M",
                                GENDER %in% c("Male") ~ "M",
                                GENDER %in% c("male") ~ "M",
                                GENDER %in% c("female") ~ "F",
                                GENDER %in% c("Female") ~ "M",
                                GENDER %in% c("Male") ~ "M",
                                GENDER %in% c("Cis Man") ~ "M",
                                GENDER %in% c("ostensibly male, unsure what that really means") ~ "M",
                                GENDER %in% c("f") ~ "F",
                                TRUE ~ "NA"
                              ) 
                       )

subset_storm_events_csv <- subset_storm_events_csv %>% mutate(
                          FAMILY_HISTORY = case_when
                              (
                                FAMILY_HISTORY %in% c("Yes") ~ "Y",
                                FAMILY_HISTORY %in% c("No") ~ "N",
                                TRUE ~ "NA"
                              ) 
                       )

subset_storm_events_csv <- subset_storm_events_csv %>% mutate(
                          TREATMENT = case_when
                              (
                                TREATMENT %in% c("Yes") ~ "Y",
                                TREATMENT %in% c("No") ~ "N",
                                TRUE ~ "NA"
                              ) 
                       )

subset_storm_events_csv <- subset_storm_events_csv %>% mutate(
                          NO_EMPLOYEES = case_when
                              (
                                NO_EMPLOYEES %in% c("More than 1000") ~ ">1000",
                                NO_EMPLOYEES %in% c("5-Jan") ~ "5-10",
                                NO_EMPLOYEES %in% c("25-Jun") ~ "25-100",
                                NO_EMPLOYEES %in% c("100-500") ~ "100-500",
                                NO_EMPLOYEES %in% c("26-100") ~ "26-100",
                                NO_EMPLOYEES %in% c("500-1000") ~ "500-1000",
                                TRUE ~ "NA"
                              ) 
                       )

subset_storm_events_csv <- subset_storm_events_csv %>% mutate(
                          REMOTE_WORK = case_when
                              (
                                REMOTE_WORK %in% c("Yes") ~ "Y",
                                REMOTE_WORK %in% c("No") ~ "N",
                                TRUE ~ "NA"
                              ) 
                       )

subset_storm_events_csv <- subset_storm_events_csv %>% mutate(
                          TECH_COMPANY = case_when
                              (
                                TECH_COMPANY %in% c("Yes") ~ "Y",
                                TECH_COMPANY %in% c("No") ~ "N",
                                TRUE ~ "NA"
                              ) 
                       )

subset_storm_events_csv <- subset_storm_events_csv %>% mutate(
                          CARE_OPTIONS = case_when
                              (
                                CARE_OPTIONS %in% c("Yes") ~ "Y",
                                CARE_OPTIONS %in% c("No") ~ "N",
                                TRUE ~ "NA"
                              ) 
                       )

subset_storm_events_csv <- subset_storm_events_csv %>% mutate(
                          WELLNESS_PROGRAM = case_when
                              (
                                WELLNESS_PROGRAM %in% c("Yes") ~ "Y",
                                WELLNESS_PROGRAM %in% c("No") ~ "N",
                                TRUE ~ "NA"
                              ) 
                       )

subset_storm_events_csv <- subset_storm_events_csv %>% mutate(
                          LEAVE = case_when
                              (
                                LEAVE %in% c("Somewhat easy") ~ "Easy",
                                LEAVE %in% c("Somewhat difficult") ~ "Difficult",
                                LEAVE %in% c("Very difficult") ~ "Difficult",
                                LEAVE %in% c("Very easy") ~ "Easy",
                                TRUE ~ "NA"
                              ) 
                       )

subset_storm_events_csv <- subset_storm_events_csv %>% mutate(
                          COWORKERS = case_when
                              (
                                COWORKERS %in% c("Some of them") ~ "Y",
                                COWORKERS %in% c("Yes") ~ "Y",
                                COWORKERS %in% c("No") ~ "N",
                                COWORKERS %in% c("1") ~ "Y",
                                COWORKERS %in% c("2") ~ "Y",
                                COWORKERS %in% c("3") ~ "Y",
                                TRUE ~ "NA"
                              ) 
                       )

subset_storm_events_csv <- subset_storm_events_csv %>% mutate(
                          SUPERVISOR = case_when
                              (
                                SUPERVISOR %in% c("Some of them") ~ "Y",
                                SUPERVISOR %in% c("Yes") ~ "Y",
                                SUPERVISOR %in% c("No") ~ "N",
                                TRUE ~ "NA"
                              ) 
                       )

print(subset_storm_events_csv)
```

## 10.	Create the following visuals :

Set of visuals that describe the data and gives a clear understanding of statistics.

1. Below is a Bar Chart displaying Gender Counts.

```{r gender histogram, echo=TRUE}
ggplot(data = subset_storm_events_csv, aes(x = GENDER, fill = factor(GENDER))) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Bar Chart of Gender Count",
    ylab = "Count",
    fill = "Gender"
  )
```

2. Below is a Bar Chart displaying Family History Counts.

```{r family history histogram, echo=TRUE}
ggplot(data = subset_storm_events_csv, aes(x = FAMILY_HISTORY, fill = factor(FAMILY_HISTORY))) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Bar Chart of Family History Count",
    ylab = "Count",
    fill = "Family History"
  )
```

3. Below is a Histogram displaying Care Options counts.

```{r Care Options Histogram, echo=TRUE}
ggplot(data = subset_storm_events_csv, aes(x = CARE_OPTIONS, fill = factor(CARE_OPTIONS))) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Histogram representing breakdown of Care Options",
    ylab = "Count",
    fill = "Care Options"
  )
```
5. Below is a Grouped Bar Plot displaying Work Interfere by Gender.

```{r Work Interfere, echo=TRUE}
ggplot(subset_storm_events_csv) +
  geom_bar(aes(x = WORK_INTERFERE, fill = GENDER), position = "dodge") +
  xlab("work_interfere") + ylab("Count")
```
6. Below is a Stacked Bar Plot displaying Work Wellness Programs by Gender.

```{r Wellness Program, echo=TRUE}
ggplot(subset_storm_events_csv) +
  geom_bar(aes(x = WELLNESS_PROGRAM, fill = GENDER)) +
  xlab("WELLNESS_PROGRAM") + ylab("Count")
```

